\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risk in a South African primary care clinic using environmental, clinical, and patient movement data}

### 2024-04-18 doc/paper/manuscript.tex

We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1~second intervals (\Cref{fig:floor-plan}). The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. \supp~Fig.~\zref{fig:tracking-examples} shows some sample tracks from our study. Note that people in the clinic could contribute multiple tracks if they moved out of the range of a sensor or if they were briefly unrecognized as a person. To address these issues, we created an R Shiny tool to manually reconnect interrupted tracks back together (\supp~Text~\zref{sec:setting-and-data}). We also labeled the tracks to identify movements of the clinic staff.

@@

(1)~We matched clinical and patient movement data to identify the spatiotemporal location of infectious TB patients within the clinic. We recorded the timestamp when the patient movement was detected in the registration area and matched this with the registration time in the clinical records. We excluded patient movements of less than five minutes and with less than five seconds in the registration area from the matching. Data were matched if there was a patient movement in the registration area followed by a entry in the clinical database, considering a maximum delay of 15 minutes. All diagnosed TB patients could be matched with patient movements this way. (2)~In addition to diagnosed TB patients, we also consider a number of undiagnosed (subclinical) TB patients. Prior studies suggest that the number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}. Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Fig.~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four TB patients visiting the clinic. Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious. We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution considering two different activity levels (\supp~Fig.~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}. Since mask wearing was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.

@@

(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to estimate air change rates by daytime using a transient mass balance model\cite{Batterman2017IJERPH}. (4)~The air change rate was related to the diffusion of quanta using, as an approximation, the empirical relationship between the eddy diffusion coefficient and the air change rate in CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}. Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated (\supp~Fig.~\zref{fig:toy-example}).

@@

(5)~Quanta removal occurs through replacement of contaminated indoor air with fresh outdoor air, inactivation of bacteria in airborne particles, and gravitational settling. The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate. The bacterial inactivation rate of \emph{Mtb} was modeled using a lognormal prior distribution (\supp~Fig.~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}. We set the gravitational settling rate to zero, thereby ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.

@@

During one week corresponding to five open clinic days, 894 patients were registered in the clinical database, of whom four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening (\supp~Fig.~\zref{fig:diagnosed-tb-patients}). Overall, 1,563 unique movements were detected in the clinic after processing the video sensor data, of which 125 were labeled as healthcare worker, identifying 1,438 clinic attendees (patients or other visitors). Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b). Clinic attendees spent about half an hour in the clinic (median 25\,min, IQR 13\,min$-$46\,min), most of it in the waiting room (\Cref{fig:input-data-descriptives}c). Close-contact encounters (proximity <1\,m and duration >1\,min) were frequent (median 5, IQR 2$-$9) and clinic attendees were in close contact with at least one other person for more than half of their visit (median 68\%, IQR 41\%$-$87\%). The CO$_2$ levels in the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\supp~Fig.~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 29 air changes per hour (IQR 9\,h$^{-1}$$-$36\,h$^{-1}$) in the densely occupied waiting room.

\begin{figure}
    \centering
     \includegraphics{results/data/no-people-spatial.png}
    \includegraphics{results/data/no-people-over-time.png}
    \includegraphics{results/data/time-in-clinic.png}
    \includegraphics{results/data/air-changes-per-hour.png}
    \caption{\textbf{Person movement characteristics and environmental conditions.} \textbf{(a)}~Total number of tracks recorded per unit space in the clinic over the study period. \textbf{(b)}~Number of people in the clinic over time per day. \textbf{(c)}~Time spent in the clinic per attendee. \textbf{(d)}~Air change rates per room of the clinic.}
    \label{fig:input-data-descriptives}
\end{figure}

\supp~File~1 shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room. It can be seen that the quanta concentration was higher near the infectious patient, indicating his position and movement. The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open during clinic hours. \Cref{fig:main-modeling-results} shows the results from our main spatiotemporal modeling analysis across study days and simulations. On average, the quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}a). An exposure of one hour to the mean quanta concentration in the middle of the waiting room would correspond to a 0.21\% risk of infection in the morning compared to a 0.05\% risk in the afternoon. The mean risk of infection per visit was 0.06\% (80\%-CrI 0.01\%$-$0.07\%), corresponding to a risk of 0.12\% per hour (80\%-CrI 0.01\%$-$0.23\%). To put this into perspective, a patient with twelve hourly visits per year would have a cumulative annual risk of infection of 1.37\% (80\%-CrI 0.14\%$-$2.74\%).

Our modeling approach allowed us to estimate the individual risk of infection with uncertainty, for example, to identify patients with a comparably higher mean risk of infection and low uncertainty (\supp~Fig.\zref{fig:main-modeling-results}). The modeled individual risk of infection was further associated with the number of close contacts and the time spent in the clinic. When the number of close contacts doubled, the odds of infection increased to 1.17 (95\%-CrI 1.04$-$1.30). Similarly, doubling the time spent in the clinic increased the odds of infection to 1.09 (95\%-CrI 1.01$-$1.18). The time spent in close contact was not significantly associated with the individual risk of infection (1.05, 95\%-CrI, 0.96$-$1.14). Finally, a comparison of our spatiotemporal modeling results with those of a temporal model assuming a well-mixed airspace showed a notable difference in the individual risk of infection with a median of 8\% (IQR 4\%$-$15\%).

\begin{figure}
    \centering
    \includegraphics{results/modeling/main-figure.png}
    \caption{\textbf{Spatiotemporal modeling results.} \textbf{(a)}~Modeled average quanta concentration in the morning and afternoon in the clinic. \textbf{(b)}~Mean risk of infection per clinic attendee per visit (median time 25\,min, IQR 13\,min$-$46\,min). Low: <5\%, Medium: 5-10\%. \textbf{(c)}~Individual mean risk of infection vs uncertainty (IQR in percentage points, pp). \textbf{(d)}~Association of individual mean risk of infection with patient movement characteristics (mean odds ratio [OR] and 95\%-credible interval [CrI] if the value of the predictor doubles).}
    \label{fig:main-modeling-results}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics{results/modeling/mean-roi-comparison.png}
    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per visit: \textbf{(a)}~with mask use and high ventilation as experienced during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as experienced before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation. Low: <5\%, Medium: 5-10\%, High: >10\% risk of infection.}
    \label{fig:scenario-results}
\end{figure}

>>>>

We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1~second intervals (\Cref{fig:floor-plan}). The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. \supp~Figure~\zref{fig:tracking-examples} shows some sample tracks from our study. Note that people in the clinic could contribute multiple tracks if they moved out of the range of a sensor or if they were briefly unrecognized as a person. To address these issues, we created an R Shiny tool to manually reconnect interrupted tracks back together (\supp~Text~\zref{sec:setting-and-data}). We also labeled the tracks to identify movements of the clinic staff.

@@

(1)~We matched clinical and patient movement data to identify the spatiotemporal location of infectious TB patients within the clinic. We recorded the timestamp when the patient movement was detected in the registration area and matched this with the registration time in the clinical records. We excluded patient movements of less than five minutes and with less than five seconds in the registration area from the matching. Data were matched if there was a patient movement in the registration area followed by a entry in the clinical database, considering a maximum delay of 15 minutes. All diagnosed TB patients could be matched with patient movements this way. (2)~In addition to diagnosed TB patients, we also consider a number of undiagnosed (subclinical) TB patients. Prior studies suggest that the number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}. Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Figure~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four TB patients visiting the clinic. Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious. We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution considering two different activity levels (\supp~Figure~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}. Since mask wearing was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.

@@

(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to estimate air change rates by daytime using a transient mass balance model\cite{Batterman2017IJERPH}. (4)~The air change rate was related to the diffusion of quanta using, as an approximation, the empirical relationship between the eddy diffusion coefficient and the air change rate in CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}. Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated (\supp~Figure~\zref{fig:toy-example}).

@@

(5)~Quanta removal occurs through replacement of contaminated indoor air with fresh outdoor air, inactivation of bacteria in airborne particles, and gravitational settling. The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate. The bacterial inactivation rate of \emph{Mtb} was modeled using a lognormal prior distribution (\supp~Figure~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}. We set the gravitational settling rate to zero, thereby ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.

@@

During one week corresponding to five open clinic days, 894 patients were registered in the clinical database, of whom four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening (\supp~Figure~\zref{fig:diagnosed-tb-patients}). Overall, 1,563 unique movements were detected in the clinic after processing the video sensor data, of which 125 were labeled as healthcare worker, identifying 1,438 clinic attendees (patients or other visitors). Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b). Clinic attendees spent about half an hour in the clinic (median 25\,min, IQR 13\,min$-$46\,min), most of it in the waiting room (\Cref{fig:input-data-descriptives}c). Close-contact encounters (proximity <1\,m and duration >1\,min) were frequent (median 5, IQR 2$-$9) and clinic attendees were in close contact with at least one other person for more than half of their visit (median 68\%, IQR 41\%$-$87\%). The CO$_2$ levels in the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\supp~Figure~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 29 air changes per hour (IQR 9\,h$^{-1}$$-$36\,h$^{-1}$) in the densely occupied waiting room.

\begin{figure}
    \centering
     \includegraphics{results/data/no-people-spatial.png}
    \includegraphics{results/data/no-people-over-time.png}
    \includegraphics{results/data/time-in-clinic.png}
    \includegraphics{results/data/air-changes-per-hour.png}
    \caption{\textbf{Patient movements and environmental conditions in the clinic.} Across clinic days and for the waiting room, corridor and TB room of the clinic: \textbf{(a)}~total number of tracks recorded per unit space, \textbf{(b)}~number of people in the clinic over time, \textbf{(c)}~time per clinic visit, and \textbf{(d)}~air change rates. Lines and ribbons show the mean plus/minus standard deviation, respectively. Boxplots show the median as lines, IQR as boxes, range as whiskers, and outliers as dots.}
    \label{fig:input-data-descriptives}
\end{figure}

\supp~File~1 shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room. It can be seen that the quanta concentration was higher near the infectious patient, indicating his position and movement. The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open during clinic hours. Across clinic days, the quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}). An exposure of one hour to the mean quanta concentration in the middle of the waiting room would correspond to a 0.21\% risk of infection in the morning compared to a 0.05\% risk in the afternoon. The mean risk of infection per clinic visit was 0.06\% (80\%-CrI 0.01\%$-$0.07\%), corresponding to a risk of 0.12\% per hour (80\%-CrI 0.01\%$-$0.23\%). To put this into perspective, a patient with twelve hourly visits per year would have a cumulative annual risk of infection of 1.37\% (80\%-CrI 0.14\%$-$2.74\%).

Our modeling approach allowed us to estimate the individual risk of infection with uncertainty, for example, to identify patients with a comparably higher mean risk of infection and low uncertainty (\supp~Figure\zref{fig:patient-risk-focus}). The modeled individual risk of infection was further associated with the number of close contacts and the time spent in the clinic. When the number of close contacts doubled, the odds of infection increased to 1.17 (95\%-CrI 1.04$-$1.30). Similarly, doubling the time spent in the clinic increased the odds of infection to 1.09 (95\%-CrI 1.01$-$1.18). The time spent in close contact was not significantly associated with the individual risk of infection (1.05, 95\%-CrI, 0.96$-$1.14). Finally, a comparison of our spatiotemporal modeling results with those of a temporal model assuming a well-mixed airspace showed a notable difference in the individual risk of infection with a median of 8\% (IQR 4\%$-$15\%).

\begin{figure}
    \centering
    \includegraphics{results/modeling/mean-quanta-concentration.png}
    \caption{\textbf{Spatiotemporal quanta concentration in the clinic.} Model-estimated average quanta concentration across clinic days in the morning and afternoon in the waiting room, corridor and TB room of the clinic.}
    \label{fig:main-modeling-results}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics{results/modeling/mean-roi-comparison.png}
    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per clinic visit: \textbf{(a)}~with mask use and high ventilation as experienced during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as experienced before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation. Low: <5\%, Medium: 5-10\%, High: >10\% risk of infection.}
    \label{fig:scenario-results}
\end{figure}

### 2024-04-18 doc/paper/supplements.tex

\begin{figure}[!htpb]
    \centering
    \includegraphics{results/data/co2-levels-over-time.png}
    \caption{CO$_2$ levels per room of the clinic over time per day.}
    \label{fig:co2-levels}
\end{figure}

\begin{figure}[!htpb]
    \centering
    \includegraphics{results/modeling/mean-roi-assumptions.png}
    \caption{\textbf{Scenario modeling results.} Mean risk of infection per clinical attendee if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic. Low: <5\%, Medium: 5-10\%, High: >10\% risk of infection.}
    \label{fig:assumptions-results}
\end{figure}

>>>>

\begin{figure}[!htpb]
    \centering
    \includegraphics{results/data/co2-levels-over-time.png}
    \caption{CO$_2$ levels per clinic room and day over time.}
    \label{fig:co2-levels}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{results/modeling/patient-risk-focus-plot.png}
    \caption{Mean and 80\%-CrI (uncertainty) of the risk of infection per clinic visit. Highlighted in red is a visit from a clinic attendee with moderate risk of infection and low estimated uncertainty.}
    \label{fig:patient-risk-focus}
\end{figure}

\begin{figure}[!htpb]
    \centering
    \includegraphics{results/modeling/mean-roi-assumptions.png}
    \caption{Mean risk of infection per clinic visit if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic. Low: <5\%, Medium: 5-10\%, High: >10\% risk of infection.}
    \label{fig:assumptions-results}
\end{figure}
